Multi-dimensional Urban Network Percolation

### 2019-09-11 Docs/Papers/Percolation/paper/Response_to_Reviewers.tex

$\rightarrow$ The study of fractal dimension is not essential in the methodology, but comes as an auxiliary result to validate the multi-dimensional approach regarding the original papers studying urban percolation. This was specified, and the formulas were better explained.

\medskip

\item \textit{A summary table would have been great here.}

$\rightarrow$ We chose not to add a summary table for fractal dimensions, as it is not the main purpose of the study and their variation is furthermore not significant.

@@

\medskip

\item \textit{Extracting endogenous megacity regions: Why is this configuration working? The impression is that the result comes as the work is felt the way along.}

$\rightarrow$ This work is exploratory, thus 4800 parameter configuration were explored. The ones shown in Fig. 4 are close to actual functional regions for some, but very far for others. This suggest that our parametrizable approach is flexible to consider varying definitions of urban regions, and that in some cases it coincides with the purely functional definition of (Hall and Pain, 2016), i.e. that some thresholds of population and road network performance intensity capture actual functional linkages. There is however no way to explain this emerging property, as our method constructs clusters from the bottom-up. This was added in the discussion.

@@

\medskip

\item \textit{How do we know that it is not just an effect of the population density layer?}

$\rightarrow$ The shape of percolation transition curves in Fig. 2 show that the addition of the road network layer adds some information, as the curve for $\theta_N = 0$ (population only) can be compared with other values. This was made more explicit.

@@

\medskip

\item \textit{``we observe a Pareto front for all points\ldots'': What do we learn about urban system from that?}

$\rightarrow$ The presence of Pareto fronts confirms that urban systems consist in general of compromises between multiple objectives. This was clarified.

### 2019-09-11 Docs/Papers/Percolation/paper/paper.tex
paper_v2

\abstract{Network percolation has recently been proposed as a method to characterize the global structure of an urban system form the bottom-up. This paper proposes to extend urban network percolation in a multi-dimensional way, to take into account both urban form (spatial distribution of population) and urban functions (here as properties of transportation networks). The method is applied to the European urban system to reconstruct endogenous urban regions. The variable parametrization allows to consider patterns of optimization for two stylized contradictory sustainability indicators (economic performance and greenhouse gases emissions). This suggests a customizable spatial design of policies to develop sustainable territories.}

@@

% TODO English was reworked throughout the paper to make it more readable and understandable.

@@

The structure of road networks can be used as a proxy to understand its past growth dynamics, but also has a significant impact on the future sustainability of territories it irrigates. Diverse methods to characterize the structure of spatial networks, and more particularly road networks, have been developed in that context, including classical network indicators such as centralities \citep{crucitti2006centrality} but also more elaborated constructions capturing more realistic processes in terms of street network use \citep{lagesse2015spatial}. Such studies of urban networks are by essence interdisciplinary, or at least imply complementary viewpoints from disciplines as diverse as architecture with space syntax \citep{hillier1976space}, physics with the study of spatial networks \citep{barthelemy2011spatial}, or social science disciplines concerned with space such as geography \citep{ducruet2014spatial}.

A method to characterize topologies of such urban spatial networks is network percolation, initially applied to road networks by \cite{arcaute2016cities}. Percolation in physics can be understood in a broad sense as processes related to the progressive occupation or connection of nodes of a network, and is generally associated to a phase transition with the emergence of a giant cluster at a given connection probability \citep{stauffer2014introduction}. Important applications include the quantification of network robustness \citep{callaway2000network} or the modeling of epidemic spreading \citep{newman1999scaling}.

Such approaches have been applied to urban systems not only for the study of networks. \cite{makse1998modeling} model urban growth with a local percolation model for site occupancy. \cite{arcaute2016cities} focus on the analysis of street networks and extract endogenous urban regions for UK which correlate with socio-economic properties, and provide a definition of urban areas which highly correlates with land-cover data. \cite{piovani2017urban} apply road network percolation at the mesoscopic scale of London metropolitan area, in relation with a retail location model. At a larger scale, the paradigm of percolation transition has been applied to the study of urban traffic dynamics \citep{Li669,Zeng23}. In spatial statistics, this method can be used to characterize the spatial morphology of point patterns \citep{huynh2018characterisation}.

Existing heuristics however generally focus on a single dimension or property of the urban system. However, such systems are known to be multidimensional, for example through the superposition of the morphological dimension of networks, and the functional properties of the urban environment \citep{burger2012form}. The link between urban form and function remains in particular an open question \citep{batty1994fractal}, but more generally the inclusion of multiple dimension in urban analysis is still a research direction to be investigated, as in the case of agent-based models for example \citep{perez2016agent}. This paper addresses such a gap in the case of urban network percolation, by introducing a multi-dimensional percolation heuristic. The method allows to combine different dimensions of the urban system, the same way that \cite{cottineau2018defining} combines population density and commuting flows to produce multiple definitions of urban areas.

@@

Beside these methodological issues of characterizing urban networks and more particularly their endogenous hierarchical structure, some related applied research issues can be considered. Indeed, quantitative tools are needed to evaluate the sustainability of recently emerged urban forms. In particular, according to \cite{lenechet2017peupler}, the most recent transition of human settlement systems (in the sense of \cite{sanders2017peupler}, i.e. a change in the dynamical regime ruling the evolution of the spatial structure of settlements) is the emergence of mega-city regions. These have been defined by \cite{hall2006polycentric} as polycentric urban structures highly integrated in terms of flows. The transition imply complex processes such as changes in the governance structure, and can not be associated to the stylized transition identified by \cite{louf2013modeling} in a simplistic toy urban model, and therefore has chances to imply more drivers than negative externalities of congestion only.

@@

Beside these targeted studies, a remaining open issue, to the best of our knowledge never tackled at this small geographical scale, is the endogenous characterization of such urban regions. The delineation of such geographical systems is taken as exogenous, and their performance and sustainability is then evaluated.

@@

% research question
% TODO clear formulation of research question : include territory - network link

This work proposes to partly tackle the two above research questions by linking them. More precisely, it first from a methodological viewpoint investigates how urban network percolation can be generalized to multiple dimensions, and secondly explores the endogenous characterization of mega-city regions using such a multi-dimensional percolation.
%  It consists in two main axis: (i) from the methodological viewpoint, how can urban network percolation be extended to multiple dimensions; and (ii) from the thematic viewpoint, how can endogenous mega urban regions be characterized which such a method, and how can it provide insights into their sustainability. These two axis are separated, but still tightly related as the second depends on the new methodology, and the first is illustrated through its concrete application in the second.

In order to test our multi-dimensional percolation method, we propose to apply it to the endogenous characterization of urban regions, and compute stylized sustainability indicators on the constructed regions.

We apply the heuristic to urban morphology and road network topology measures in Europe. The idea to combine urban form with network topology measures relies on the capture of the link between urban form and function as already mentioned, urban functions being assumed as distributed by transportation networks \citep{raimbault2018caracterisation}.

% TODO dimension of the urban system taken into account were better specified in the introduction

Our contribution relies on several points: (i) this is to the best of our knowledge the first time a multi-dimensional percolation method is applied to urban systems; (ii) we furthermore apply it on the significant spatial extent of all European Union; and (iii) we link the clusters obtained with simple sustainability measures. The rest of this paper is organized as follows: we first describe the multi-dimensional percolation heuristic, the data and variables to which it is applied, and the indicators used to characterized the sustainability of clusters produced. We then describe the results of applying this method to population and network variables for the whole European Union, focusing on the endogenous regions produced and their sustainability properties. We finally discuss possible developments and the implications of this methodology to the design of policies.

% TODO add more details on all aspects of methodology

% TODO more details in the text on formulas and parameters

@@

% TODO rewrite legend of fig 1

%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[ht] 
%\resizebox{10cm}{7cm}
  {\includegraphics[width=\linewidth]{figures/principle.pdf}}
  \centering
  \caption{Schematic representation of the multi-dimensional network percolation heuristic. We show a stylized configuration with two layers having the same nodes, and the links within each layer are created following the percolation radius $r_0$ and the thresholds for the layer variables. The final clusters are the superposition of these. The green points give examples of starting points in the case of a propagation heuristic.\label{fig:method}}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

@@

Note that we do not call our method ``multi-layer percolation'', as it is not strictly multi-layer since nodes are common. The term of multi-dimensional percolation is more suited to the use of multiple variables and thresholds. The method works with an arbitrary number of layers. See Fig.~\ref{fig:method} for a schematic representation of the method. It can be implemented with a propagation heuristic or directly working on adjacency matrices. The rationale behind the conjunction of the thresholding of each layer variable and the distance thresholding relies on the idea that two points will interact if they are close enough, but also if they have a strong enough intensity of the activity or dimension captured by each layer, simultaneously for all layers considered. This recalls Tobler's first law of geography \citep{tobler2004first} in a multi-dimensional way.

@@

More formally, let assume a set of nodes $V = v_i$ common to all layers of the network, and layers edges $E_j$ for layer $j$ taken as empty at the initial state of the algorithm. Each node has a value of the considered variables associated to each layer, written $v_{ij}$. For each layer, a link $e_{kl} \in E_j$ is created if $d(v_k,v_l) < r_0$ where $d$ is the distance between the nodes (which can be any distance) and $r_0$ the percolation radius, and if $v_{kj} > \theta_j$ and $v_{lj} > \theta_j$ where $\theta_j$ is the threshold for layer $j$. The final percolated network edges $E$ is composed by links contained within all layers simultaneously. The multi-dimensional percolation clusters are then the connected components of this network $(V,E)$. The parameters implied in this heuristic are the percolation radius $r_0$ and the percolation thresholds $\theta_j$ for each layer $j$, allowing a flexible application through parametrization.

@@

The previous generic method must be applied to a consistent urban network, in the sense of multiple variables associated to spatial nodes. To each variable then corresponds a network layer. We propose to work on gridded networks, namely nodes regularly spaced in two dimensions. In practice, these nodes will be the center of raster cells. We consider two layers, one defined by population density and the other by road network structure indicators computed at the same resolution.
% TODO French-like paragraph

\subsection{Empirical data}

@@

We use this data to construct a two layers abstract network: a layer which variable is given by population density, and a second layer which variable is given by a road network indicator. Nodes are the center of cells (thus disposed in space on a grid of step 5km). We test several possible networks by varying the road indicator taken into account for the second layer. In particular, we test the following indicators:
\begin{enumerate}
	\item number of edges $N_E$;
	\item number of vertices $N_V$;
	\item cyclomatic number $\mu$ which is defined by $\mu = N_E - N_V + c$ where $c$ is the number of connected components of the graph; this measure captures the number of distinct cycles in the graph
	\item euclidian efficiency $v_0$, defined by \cite{banos2012towards}
\end{enumerate}

These measures capture functional properties especially for the two last.

% TODO The choice of network indicators was further discussed, in relation with the research question.

% TODO remove the term network variable ? network indicators

The percolation on such an abstract network is a necessary condition in our case to link the different dimensions considered, namely population distribution and local road network properties. We have therefore two levels of networks in our approach, namely the physical road network which local properties are taken here as input, and the abstract two layer network on which we do the percolation. We will in the following write $\theta_P$ for the threshold parameter of the population layer, and $\theta_N$ for the threshold parameter of the network layer. In practice, these parameters will be given in the following as quantile level of the corresponding variable, for an easier interpretation and conception of experience plans. The name of the variable considered will be written $v_N$.

@@

As already detailed, urban regions may perform more or less well regarding different dimensions of sustainability. We propose to use the endogenous definition of regional urban systems produced by the percolation algorithm to evaluate their sustainability, in terms of conflicting objectives of economic integration and greenhouse gases emissions. The definition of sustainability, or sustainable development, is by essence multi-dimensional \citep{viguie2012trade}. Its characterization as quantitative indicators is even more subject to numerous degrees of freedom. We work here with two stylized indicators for two conflicting dimensions, as a proof-of-concept.

% TODO The choice of these dimensions of sustainability is to give an illustration with two paradigmatic opposed dimensions, which can furthermore be approximated indirectly from gravity models as we propose. The explanation of this choice was added in the introduction.

@@

Applying a gravity model to each region, we estimate abstract transportation flows within each and use these to extrapolate emissions from the actual local emission from the Edgar database, and economic activities with a scaling law of population.  More precisely, following \cite{raimbault2018indirect}, a potential flow between two points $i$ and $j$ can be estimated with the following expression

@@

where $v^{(k)}_i$ are either population or effective local GHG emissions (indexed by $k = 1,2$ respectively), $d_{ij}$ the distance between the two points, $d_0$ a distance decay parameter, and $\gamma$ a scaling parameter. Indeed, the economic activity follows relatively well scaling laws of populations \citep{bettencourt2007growth}, the exponent being dependant on the activity and the definition of areas on which it is estimated~\citep{cottineau2017diverse}.

The sum of all flows of points within the geographical span of the cluster (that we approximate as the convex Hull envelope of its points), allows us to approximate the cumulated potential emissions and economic activity. Writing clusters $K_c$ as this set of points, we define the total economic flow by

@@

This allows defining a relative economic inefficiency by

@@

and relative potential emissions by

@@

Both indicators should be minimized for sustainability. Normalized indicators $\tilde{e}_c,\tilde{g}_c$ are defined in a similar way, but the extrema being computed on all other possible urban configurations with the same $\gamma,d_0$ values.

@@

@@

We recall that the euclidian performance of the network \citep{banos2012towards} is in our case $<d_e/d_n>$ where the average is taken on all origin-destination pairs in the network, $d_e$ is the euclidian distance and $d_n$ the network distance. Thus, it indeed increases with network performance, in consistence with the use done here through thresholding.
% TODO clarify

@@

In its application to road networks by \cite{arcaute2016cities}, the structure of the national urban system for UK is captured by studying the percolation transition, i.e. the variation of the size of the largest cluster as a function of the percolation radius. As this signature is tightly linked to historical, cultural and geographical conditions, the application to different urban systems should yield different results. We study here this property, for different threshold parameter values. We make the radius vary betweem 8km and 100km with a one km step, have a fixed population threshold $\theta_P = 0.85$, test all network variables, and three network thresholds $\theta_N \in \{ 0 ; 0.8 ; 0.95 \}$.

The absolute and relative sizes of the largest cluster are plotted in Fig.~\ref{fig:percolation} as a function of the percolation radius. This aspect first gives methodological information on multilayer percolation. Indeed, comparing the result with $\theta_N = 0$ (single layer percolation) with positive values of $\theta_N$ shows a significantly different behavior. As expected, absolute size are much smaller, but when looking at relative sizes we observe that the abrupt steps typical to percolation transitions have different distributions across the different parametrizations. The more regular curve seems to be the standard percolation on population only, whereas at $\theta_N = 0.95$, different network variables produce either very early transitions (for $\mu$ for example) or very late (for $N_V$). Also, changing of scale compared to \cite{arcaute2016cities} gives more steps and less abrupts curves in general, confirming the integration of subsystems with different structures in our analysis and the importance of scale in such analysis. As the addition of a layer also changes drastically the results, one should stay careful when switching from a mono-dimensional percolation to a multi-dimensional percolation.

@@

We study also the evolution of the fractal dimension of clusters as a function of $r_0$. Following \cite{arcaute2016cities}, we estimate the fractal dimension of clusters $\alpha$ with a simple OLS regression between cluster size and cluster diameter, namely $\log N_c = k + \alpha \cdot \log \delta_c$ where $N_c$ is the size of cluster $c$ and $\delta_c$ its diameter. As a negative result, that could be due to the abstract nature of our network, a clear maximum in the value of the fractal dimension can not be found. Either it is located at resolution that our method can not reached due to the minimal 5km limit imposed by the abstraction, or it does not exist when coupling dimensions. Determining which assumption is more plausible is out of the scope of this paper. We do not plot the standard error $\sigma$ of fractal dimensions for visibility purposes, but their relative value given by $\alpha / \sigma\left[\alpha\right]$ is in average 0.10 and in maximum 0.196 on all points, meaning that these estimations remain however consistent. Regarding the variability as a function of the percolation radius $r_0$, studying the difference $(\alpha - \sigma\left[\alpha\right])_M - (\alpha - \sigma\left[\alpha\right])_m$ where the first is taken at maximum for $\alpha$ and the other at minimum, shows that the configuration for $\mu$ and $\theta_N=0.95$ has a clearly significant maximum (difference at 0.38). For this coupling, the endogenous structure given by the maximum may be defined. Other configurations yield non-significant maximums.

% TODO The study of fractal dimension is not essential in the methodology, but comes as an auxiliary result to validate the multi-dimensional approach regarding the original papers studying urban percolation. This was specified, and the formulas were better explained.

This study of percolation transition and fractal dimension shows that our multi-dimensional percolation heuristic remains relevant, as results analog to the one-dimensional approach can be obtained.

% TODO table ?
%  fractdimsignif
%# A tibble: 11 x 5
%# Groups:   nwcol, popthq, nwthq [?]
%   nwcol    popthq nwthq nwparam       signif
%   <fct>     <dbl> <dbl> <chr>          <dbl>
% 1 ecount     0.85  0    ecount0      -0.105 
% 2 ecount     0.85  0.8  ecount0.8    -0.0452
% 3 euclPerf   0.85  0    euclPerf0    -0.108 
% 4 euclPerf   0.85  0.8  euclPerf0.8   0.0360
% 5 euclPerf   0.85  0.95 euclPerf0.95 -0.108 
% 6 mu         0.85  0    mu0          -0.108 
% 7 mu         0.85  0.8  mu0.8        -0.108 
% 8 mu         0.85  0.95 mu0.95        0.375 
% 9 vcount     0.85  0    vcount0      -0.105 
%10 vcount     0.85  0.8  vcount0.8    -0.0416
%11 vcount     0.85  0.95 vcount0.95   -0.0868

@@

% TODO This part was clarified by adding more context literature and more explanations.

@@

% TODO extend with relation to research questions

% TODO But the only fact of applying a new methodology does not make a research result to me: it is necessary to show how brought information is significant to analyze an urban system

% TODO The discussion was extended in relation with answers to the research question.

>>>>

\abstract{Network percolation has recently been proposed as a method to characterize the hierarchical structure of an urban system form the bottom-up. This paper proposes to extend urban network percolation in a multi-dimensional way, to take into account both urban form (spatial distribution of population) and urban functions (here as properties of transportation networks). The method is applied to the European urban system to reconstruct endogenous urban regions. The variable parametrization allows to consider patterns of optimization for two stylized contradictory sustainability indicators (economic performance and greenhouse gases emissions). This suggests a customizable spatial design of policies to develop sustainable territories.}

@@

@@

The structure of road networks can be used as a proxy to understand its past growth dynamics, but also has a significant impact on the future sustainability of territories it irrigates. Diverse methods to characterize the structure of spatial networks, and more particularly road networks, have been developed in that context. They include classical network indicators such as centralities \citep{crucitti2006centrality} but also more elaborated constructions capturing more realistic processes in terms of street network use \citep{lagesse2015spatial}. Such studies of urban networks are by essence interdisciplinary, or at least imply complementary viewpoints from diverse disciplines. These for example include architecture with space syntax \citep{hillier1976space}, physics with the study of spatial networks \citep{barthelemy2011spatial}, or social science disciplines concerned with space such as geography \citep{ducruet2014spatial}.

A method to characterize the hierarchical structure of such urban spatial networks is network percolation, initially applied to urban road networks by \cite{arcaute2016cities}. Percolation in physics can be understood in a broad sense as processes related to the progressive occupation or connection of nodes of a network. It is generally associated to a phase transition with the emergence of a giant cluster at a given probability of connection \citep{stauffer2014introduction}. Practical applications in different fields include the quantification of network robustness \citep{callaway2000network} or the modeling of epidemic spreading \citep{newman1999scaling}.

Such approaches have been applied to urban systems with other applications than the study of networks. \cite{makse1998modeling} model urban growth with a local percolation model for site occupancy. \cite{arcaute2016cities} focus on the analysis of street networks and extract endogenous urban regions for UK which correlate with socio-economic properties, and provide a definition of urban areas which highly correlates with land-cover data. \cite{piovani2017urban} apply road network percolation at the mesoscopic scale of London metropolitan area, in relation with a retail location model. At a larger scale, the paradigm of percolation transition has been applied to the study of urban traffic dynamics \citep{Li669,Zeng23}. In spatial statistics, this method can be used to characterize the spatial morphology of point patterns \citep{huynh2018characterisation}.

Existing heuristics generally focus on a single dimension or property of the urban system. However, such systems are known to be highly multidimensional. For example, the morphological dimension of networks is complementary to the functional properties of the urban environment \citep{burger2012form}. The link between urban form and function remains in particular an open question \citep{batty1994fractal}. More generally, the inclusion of multiple dimension in urban analysis is still a research direction to be investigated, as in the case of agent-based models for example \citep{perez2016agent}. This paper addresses such a gap in the case of urban network percolation, by introducing a multi-dimensional percolation heuristic. The method allows combining different dimensions of the urban system, the same way that \cite{cottineau2018defining} link population density with commuting flows to produce multiple definitions of urban areas.

@@

Beside these methodological issues of characterizing urban networks and more particularly their endogenous hierarchical structure, some related applied research issues can be considered. Indeed, quantitative tools are needed to evaluate the sustainability of recently emerged urban forms. In particular, according to \cite{lenechet2017peupler}, the most recent transition of human settlement systems (in the sense of \cite{sanders2017peupler}, i.e. a change in the dynamical regime ruling the evolution of the spatial structure of settlements) is the emergence of mega-city regions. These have been defined by \cite{hall2006polycentric} as polycentric urban structures highly integrated in terms of flows. The transition imply complex processes such as changes in the governance structure, and can not be associated to the stylized transition identified by \cite{louf2013modeling} in a toy urban model including negative externalities of congestion only.

@@

Beside these targeted studies, a remaining open issue, to the best of our knowledge never tackled at this small geographical scale, is the endogenous characterization of such urban regions. The delineation of these geographical systems is often taken as exogenous, and their performance and sustainability is then evaluated. Several approaches can be taken to define such systems, including integration through transportation networks, continuity of night lights, and economic productivity thresholds~\citep{lang2005beyond,florida2008rise}. In these existing studies, parameters to define such regions remain fixed, and an endogenous hierarchical structure is possibly ignored. The second component of our research question will thus be the endogenous characterization of urban regions and how this can be applied to study their sustainability.

@@

This work proposes to partly tackle the two above research questions by linking them. More precisely, it first investigates from a methodological viewpoint how urban network percolation can be generalized to multiple dimensions, and secondly explores the endogenous characterization of mega-city regions using such a multi-dimensional percolation, and how this can be used to quantify the sustainability of these systems. These two axis are tightly linked. Indeed, on the one hand street network percolation has initially been proposed to identify endogenous entities in territorial systems, and on the other hand mega-urban regions are characterized simultaneously by morphological dimensions (continuity of the built environment) and functional dimensions (high integration of flows).

We therefore consider percolation on two dimensions of the urban systems, one linked to urban form which is the spatial distribution of population, and one linked to transportation networks, through structural indicators quantifying local road networks. We apply in particular the heuristic to urban morphology and road network topology measures in Europe. The idea to combine urban form with network topology measures relies on the capture of the link between urban form and function as already mentioned, urban functions being assumed as distributed by transportation networks \citep{raimbault2018caracterisation}. In general, the interactions between transportation networks and territories have been shown to be central processes in urban dynamics \citep{espacegeo2014effets}.

We thus introduce in this paper a multi-dimensional percolation method, and apply it to the endogenous characterization of urban regions, to finally compute stylized sustainability indicators on the constructed regions. These sustainability measures are extrapolated from gravity flows, and are used as a proof-of-concept of how this work can be applied towards policy-making. As detailed below, we use paradigmatic opposed dimensions of sustainability which are greenhouse gases emissions and economic productivity.

The originality of our contribution relies thus on several points: (i) this is to the best of our knowledge the first time a multi-dimensional percolation method is applied to urban systems; (ii) we furthermore apply it on the significant spatial extent of all European Union; and (iii) we link the clusters obtained with simple sustainability measures.

The rest of this paper is organized as follows: we first describe the multi-dimensional percolation heuristic, the data and variables to which it is applied and how the network is constructed, and the indicators used to characterized the sustainability of clusters produced. We then describe the results of applying this method to population and road network indicators for the whole European Union, focusing on the endogenous regions produced and their sustainability properties. We finally discuss possible developments and the implications of this methodology to the design of policies.

@@

%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[ht] 
%\resizebox{10cm}{7cm}
  {\includegraphics[width=\linewidth]{figures/principle.pdf}}
  \centering
  \caption{Schematic representation of the multi-dimensional network percolation heuristic. Two layers are considered here, with similar nodes but different links in each. To each layer is associated a node variable. In each layer, links are created according to a distance threshold (percolation radius) and a variable value threshold. The final clusters (bottom in color) are obtained by considering the links present in each layer.\label{fig:method}}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

@@

We do not call our method ``multi-layer percolation'', since nodes are common between layers. We show in Fig.~\ref{fig:method} for a schematic representation of the method. It can be implemented with a propagation heuristic or directly working on adjacency matrices. The rationale behind combining a thresholding for each layer variable with a distance thresholding relies on the idea that a first component for two points to interact is a large enough proximity. The second component is a strong enough intensity of the urban dimension captured by each layer, simultaneously for all layers considered. This recalls Tobler's first law of geography \citep{tobler2004first} in a multi-dimensional way.

@@

More formally, let assume a set of nodes $V = v_i$ common to all layers of the network, and layers edges $E_j$ for layer $j$ taken as empty at the initial state of the algorithm. Each node has a value of the considered variables associated to each layer, written as a matrix $v_{ij}$. The algorithm works as follows:
\begin{enumerate}
	\item Percolation is first done within each layer. For each layer, a link $e_{kl} \in E_j$ is created if $d(v_k,v_l) < r_0$ where $d$ is the distance between the nodes (which can be any distance) and $r_0$ the percolation radius, and if $v_{kj} > \theta_j$ and $v_{lj} > \theta_j$ where $\theta_j$ is the threshold for layer $j$.
	\item Layers are combined, by computing the final percolated network edges $E$ as the links contained within all layers simultaneously. The multi-dimensional percolation clusters are then the connected components of this network $(V,E)$.
\end{enumerate}

The parameters implied in this heuristic are the percolation radius $r_0$ and the percolation thresholds $\vec{\theta} = \theta_j$ for each layer $j$. Varying these parameters allows considering different levels of percolation.

@@

The previous generic method must be applied to a consistent urban network, in the sense of multiple variables associated to spatial nodes. To each variable then corresponds a network layer. We propose to work on gridded networks, namely nodes regularly spaced in two dimensions. In practice, these nodes will be the center of raster cells. We will consider two layers, one defined by population density and the other by road network structure indicators computed at the same resolution.

\subsection{Empirical data and network construction}

@@

We use this data to construct a two layers abstract network: a layer which variable is given by population density, and a second layer which variable is given by a road network indicator. Nodes are the center of cells (thus disposed in space on a grid of step 5km). We test several possible networks by varying the road indicator taken into account for the second layer. In particular, we test the following indicators:
\begin{enumerate}
	\item number of edges $N_E$;
	\item number of vertices $N_V$;
	\item cyclomatic number $\mu$ which is defined by $\mu = N_E - N_V + c$ where $c$ is the number of connected components of the graph; this measure captures the number of distinct cycles in the graph
	\item euclidian efficiency $v_0$, defined by \cite{banos2012towards}, as the average rate between network distance and euclidian distance for all couples of links
\end{enumerate}

The choice of these measures is aimed at capturing basic aspects of network structure, and functional properties especially for the two last. Indeed, euclidian efficiency measures how the network is performant to link nodes, while the number of cycles is linked to redundancy of paths and in a way to robustness. These choice are arbitrary, but several aspects of transportation networks are still captured by these indicators. An increase of each is related to a more performant network regarding different dimensions, what is relevant for our application. A systematic study with other indicators such as centralities or accessibilities is out of the scope of this paper.

The percolation on such an abstract network is a necessary condition in our case to link the different dimensions considered, namely population distribution and local road network properties. We have therefore two levels of networks in our approach, namely the physical road network which local properties are taken here as input, and the abstract two layer network on which we do the percolation.

We will in the following write $\theta_P$ for the threshold parameter of the population layer, and $\theta_N$ for the threshold parameter of the network layer. In practice, these parameters will be given in the following as quantile level of the corresponding variable, for an easier interpretation and conception of experience plans. The name of the road network indicator considered will be written $v_N$.

@@

As already detailed, urban regions may perform more or less well regarding different dimensions of sustainability. We propose to use the endogenous definition of regional urban systems produced by the percolation algorithm to evaluate their sustainability, in terms of conflicting objectives of economic integration and greenhouse gases emissions. The definition of sustainability, or sustainable development, is by essence multi-dimensional \citep{viguie2012trade}. Its characterization as quantitative indicators is even more subject to numerous degrees of freedom. We work here with these two stylized indicators for two conflicting dimensions, as a proof-of-concept. These dimensions can furthermore be approximated indirectly from gravity models as we will describe below. By introducing other datasets, our work could be extended to more realistic indicators and other dimensions.

@@

Applying a gravity model to each region, we estimate abstract transportation flows within each with a gravity model. More precisely, following \cite{raimbault2018indirect}, a potential flow between two points $i$ and $j$ can be estimated with the following expression

@@

where $v^{(k)}_i$ are either population or effective local GHG emissions computed with the EDGAR database (indexed by $k = 1,2$ respectively), $d_{ij}$ the distance between the two points, $d_0$ a distance decay parameter, and $\gamma$ a scaling parameter. Indeed, the economic activity follows relatively well scaling laws of populations \citep{bettencourt2007growth}, the exponent being dependant on the activity and the definition of areas on which it is estimated~\citep{cottineau2017diverse}. The distance decay captures the geographical span of potential interactions. These two parameters $\gamma,d_0$ are left free and varying them allows considering stylized configurations, such as long or short span interactions, or infra- or supra-linear scaling activities. Finally, considering the flow with the population variable ($k=1$) provides a proxy for economic flows, while with the GHG emissions ($k=2$) it provides a proxy for emissions in relation with this economic activity. Indeed, effective emissions are linked to local emissions and transportation emissions linked to the intensity of flows.

We then consider the sum of all these flows of points within the geographical span of a given cluster of nodes in our network. These clusters are obtained with the percolation method described above, and are numbered by $1 \leq c \leq C$. For the sake of simplicity, we approximate the corresponding geographical area as the convex Hull envelope of the points in the cluster, that we write $K_c$. By cumulating the flows, we therefore define the total economic flow as the sum of economic flows by

@@

These indicators have no interpretable unit and need to be renormalized. This allows defining a relative economic inefficiency by

@@

where $\max_c E_c$ (resp. $\min_c E_c$) is the maximal (resp. minimal) value of $E_c$ across all clusters. We define in a similar way the relative potential emissions by

@@

where $\max_c G_c$ (resp. $\min_c G_c$) is the maximal (resp. minimal) value of $G_c$ across all clusters.

Both indicators should be minimized for sustainability. Their value is dependant on the number of clusters and their extent, i.e. the geographical surface they cover. To be able to compare values across different clusterings (corresponding to different parameter values for the percolation heuristic), we finally define normalized indicators $\tilde{e}_c,\tilde{g}_c$ in a similar way, but the extrema being computed on all other possible urban configurations with the same $\gamma,d_0$ values.

@@

For descriptive purposes, we also consider summary measures of clusters, as the population $P_c$ and effective emissions $EM_c$ taken as the sum of population (resp. emissions) of the points in $K_c$.

@@

@@

In its application to road networks by \cite{arcaute2016cities}, the structure of the national urban system for UK is captured by studying the percolation transition, i.e. the variation of the size of the largest cluster as a function of the percolation radius. As this signature is tightly linked to historical, cultural and geographical conditions, the application to different urban systems should yield different results. We study here this property, for different threshold parameter values. We make the radius vary betweem 8km and 100km with a one km step, have a fixed population threshold $\theta_P = 0.85$, test all road network indicators, and three network layer thresholds $\theta_N \in \{ 0 ; 0.8 ; 0.95 \}$.

The absolute and relative sizes of the largest cluster are plotted in Fig.~\ref{fig:percolation} as a function of the percolation radius. This aspect first gives methodological information on multilayer percolation. Indeed, comparing the result with $\theta_N = 0$ (single layer percolation) with positive values of $\theta_N$ shows a significantly different behavior. As expected, absolute size are much smaller, but when looking at relative sizes we observe that the abrupt steps typical to percolation transitions have different distributions across the different parametrizations. This result is particularly interesting regarding the first axis of our research question, as it shows that the structure of clusters obtained is not only due to the population layer, and that the multi-dimensional percolation captures a complementary signal.

The more regular curve seems to be the standard percolation on population only, whereas at $\theta_N = 0.95$, different road network indicators produce either very early transitions (for $\mu$ for example) or very late (for $N_V$). Also, changing of scale compared to \cite{arcaute2016cities} gives more steps and less abrupts curves in general, confirming the integration of subsystems with different structures in our analysis and the importance of scale in such analysis. As the addition of a layer also changes drastically the results, one should stay careful when switching from a mono-dimensional percolation to a multi-dimensional percolation.

@@

We study also the evolution of the fractal dimension of clusters as a function of $r_0$, to verify how the initial percolation approach is robust to multi-dimensionality. Following \cite{arcaute2016cities}, we estimate the fractal dimension $\alpha$ of clusters with a simple OLS regression between cluster size and cluster diameter, namely

\begin{equation}
\log N_c = k + \alpha \cdot \log \delta_c
\end{equation}

where $N_c$ is the size of cluster $c$ and $\delta_c$ its diameter. These estimations are shown in Fig.~\ref{fig:fractaldim}. As a negative result, which could be due to the abstract nature of our network, a clear maximum in the value of the fractal dimension can not be found. Either it is located at a resolution that our method can not reach due to the minimal 5km limit imposed by the abstraction in the network construction, or it does not exist when coupling dimensions. Determining which assumption is more plausible is out of the scope of this paper.

We do not plot the standard error $\sigma$ of fractal dimensions (obtained as the estimation error in the OLS) for visibility purposes, but their relative value given by $\alpha / \sigma\left[\alpha\right]$ is in average 0.10 and in maximum 0.196 on all points, meaning that these estimations remain however consistent.

Regarding the variability of fractal dimension as a function of the percolation radius $r_0$, we study the possible existence of a significant maximum when $r_0$ varies. We therefore simply consider the difference $(\alpha - \sigma\left[\alpha\right])_M - (\alpha - \sigma\left[\alpha\right])_m$ where the first is taken at the maximum value of $\alpha$ and the other at its minimum value. This intuitively corresponds to checking if confidence intervals do not overlap between the maximum and the minimum of the curve. We find that the configuration for $\mu$ and $\theta_N=0.95$ has a clearly significant maximum (difference at 0.38). For this coupling, the endogenous structure given by the maximum may be defined. Other configurations yield non-significant maximums (negative values of the difference).

This study of percolation transition and fractal dimension thus shows that our multi-dimensional percolation heuristic remains relevant, as results analog but qualitatively different to the one-dimensional approach can be obtained.

%  fractdimsignif
%# A tibble: 11 x 5
%# Groups:   nwcol, popthq, nwthq [?]
%   nwcol    popthq nwthq nwparam       signif
%   <fct>     <dbl> <dbl> <chr>          <dbl>
% 1 ecount     0.85  0    ecount0      -0.105 
% 2 ecount     0.85  0.8  ecount0.8    -0.0452
% 3 euclPerf   0.85  0    euclPerf0    -0.108 
% 4 euclPerf   0.85  0.8  euclPerf0.8   0.0360
% 5 euclPerf   0.85  0.95 euclPerf0.95 -0.108 
% 6 mu         0.85  0    mu0          -0.108 
% 7 mu         0.85  0.8  mu0.8        -0.108 
% 8 mu         0.85  0.95 mu0.95        0.375 
% 9 vcount     0.85  0    vcount0      -0.105 
%10 vcount     0.85  0.8  vcount0.8    -0.0416
%11 vcount     0.85  0.95 vcount0.95   -0.0868

@@

@@

\subsection{From multi-dimensional percolation to the sustainability of mega-city regions}
